---
title: "juliaent_FinalR"
output: html_document
---

```{r}
flying = read.table(file="https://raw.githubusercontent.com/OscarFHC/NRE538_2017Fall/master/Final/flying.csv",header=TRUE, sep=",")

college = read.table(file="https://raw.githubusercontent.com/OscarFHC/NRE538_2017Fall/master/Final/college.csv",header=TRUE, sep=",")

happy = read.table(file="https://raw.githubusercontent.com/OscarFHC/NRE538_2017Fall/master/Final/happy.csv",header=TRUE, sep=",")

cancer = read.table(file="https://raw.githubusercontent.com/OscarFHC/NRE538_2017Fall/master/Final/cancer.csv",header=TRUE, sep=",")
```

## 1)	Is there a significant association between gender (gender) and whether people think it?۪s rude to bring an unruly child on the plane (unruly_child)? flying

## A Chi-square test should be used due to both variables being categorical data. Assumptions include independent observations and no structural zeros. To remove the NA values, I subset the data. The results of the test yield a significant p-value, meaning that there is an association between gender and whether people think it's rude to bring an unruly child on the plane.


```{r}
library(dplyr)
data = flying
flying = flying %>%
subset(is.na(flying$gender)==FALSE & is.na(flying$unruly_child)==FALSE)
chisq.test(flying$gender, flying$unruly_child)
```

## 2)	Is there a significant difference in tuition (tuition) by type of institution (type)? college

## The IV is categorical and binary and the DV is continuous so a t.test is an appropriate test to use. The assumptions for a t.test are continuous data, a randomly selected and independent sample, normal values or the sample size is sufficiently large, and equal variances. The assumption of normal distribution is not met given the output of the shapiro test, but the sample size is sufficiently large so we will assume this is okay. The assumption of equal variances is also not met so Welch's t.test must be used. The results show that there is a significant difference in tuition by type of institution. The average tuition of private nonprofit institutions is higher than the average tuition of public institutions.


```{r}
data = college
shapiro.test(college$tuition)
var.test(college$tuition~college$type)
type.PUB = subset(college, type=="Public")
type.PRI = subset(college, type=="Private nonprofit")
t.test(type.PUB[,"tuition"], type.PRI[,"tuition"], paired=FALSE, var.equal = FALSE)
```

## 3)	Is there a significant difference in happiness (Hscore) by region (Region)? happy

## The IV is categorical, but contains more than two groups and the DV is continuous so a one-way ANOVA is the appropriate test to use. Assumptions for ANOVA are normal distribution, independent samples, and equal variances. The data does not have a normal distribution, but the sample size is sufficiently large, and variances are not equal. Results show that there is a significant difference in happiness by region. 

```{r}
library(car)
data = happy
shapiro.test(happy$Hscore)
leveneTest(Hscore~Region,data=happy)
mod = aov(happy$Hscore~happy$Region)
summary(mod)
mod1 = lm(Hscore~Region, data=happy)
summary(mod1)
mod2 = lm(Hscore~Region-1, data=happy)
summary(mod2)
TukeyHSD(mod)
```

## 4)	What factors are significantly associated with a country?۪s corruption levels (Corruption)? Choose three continuous independent variables to include in your model. happy

## The three continuous variables I chose were "Freedom" "GDP" and "Generosity" because these variables are not highly correlated with each other. Assumptions checked for the linear model are a linear relationship between variables, homoscedasticity, independence of errors, and normality of error distribution. The DW test shows a linear relationship. The BP test shows homoscedasticity. The model does not pass the Shapiro test for normality, but the qq plot looks normal so we will assume normality. Results show that Freedom is the variable most significantly associated with Corruption. GDP and Generosity have a weaker relationship with Corruption, but still significant at the .05 level.

```{r}
data = happy
cor(happy[, c("Freedom", "GDP", "Generosity", "Corruption")], use="na.or.complete")
pairs(happy[, c("Freedom", "GDP", "Generosity", "Corruption")])
mod3 = lm(Corruption~Freedom+GDP+Generosity, data=happy)
summary(mod3)
library(lmtest)
dwtest(mod3, alternative=c("two.sided"))
bptest(mod3)
plot(mod3)
qqnorm(happy$Corruption)
qqline(happy$Corruption)
shapiro.test(happy$Corruption)
plot(residuals(mod3))
shapiro.test(residuals(mod3))
vif(mod3)
aov(mod3)
```

## 5)	Choose one of the continuous independent variables that was significant in the model for Question 4 and interact it with region (Region) to predict corruption (Corruption). This model should only include one continuous independent variable and its interaction with region. Does the influence of your continuous variable on corruption vary by region? happy

## Assumptions for the ANCOVA are the same as for the linear model, linear relationship between variables, homoscedasticity, independence of errors, and normality of error distribution. The DW test shows a linear relationship. The BP test shows homoscedasticity. The model does not pass the Shapiro test for normality, but the qq plot looks normal so we will assume normality. Results show that the influence of freedom on corruption does vary by region.

```{r}
data = happy
mod4 = lm(Corruption~Freedom+Region, data=happy)
summary(mod4)
dwtest(mod4, alternative=c("two.sided"))
bptest(mod4)
qqnorm(happy$Corruption)
qqline(happy$Corruption)
shapiro.test(happy$Corruption)
plot(mod4)
plot(residuals(mod4))
shapiro.test(residuals(mod4))
vif(mod4)
aov(mod4)
```

## 6)	Which factors are significantly associated with whether a breast cancer tumor is malignant or not? Choose three continuous independent variables to include in your model. cancer

## The assumptions checked for the linear model are a linear relationship between variables, homoscedasticity, independence of errors, and normality of error distribution. The DW test shows a linear relationship. The BP test shows homoscedasticity. The model does not pass the Shapiro test, but the qq plot looks normal so we will assume normality. The error distribution is not normal so we will try a log transform of the dependent variable. All of the three factors "radius_mean", "texture_mean", and "perimeter_mean" are significantly associated with whether a breast cancer tumor is malignant or not.

```{r}
data = cancer
log(cancer$malignant)
mod5 = lm(malignant~radius_mean+texture_mean+perimeter_mean, data = cancer)
summary(mod5)
dwtest(mod5, alternative=c("two.sided"))
bptest(mod5)
qqnorm(cancer$malignant)
qqline(cancer$malignant)
shapiro.test(cancer$malignant)
plot(mod5)
plot(residuals(mod5))
shapiro.test(residuals(mod5))
vif(mod5)
aov(mod5)
```

## 7)	BONUS/EXTRA CREDIT: Which independent variables are the most important in explaining whether a breast cancer tumor is malignant or not? Use the same 3 continuous independent variables you chose for question 6. cancer.

## "Perimeter_mean" is the most important in explaining whether a breast cancer tumor is malignant or not, closely followed by "radius_mean", "texture_mean" is the least important of these three variables. 

```{r}
cor(cancer[, c("malignant", "radius_mean", "texture_mean", "perimeter_mean")], use="na.or.complete")
```

