---
title: "EAS 538 Generalized Linear Model : Answer Key"
author: "Written by Oscar Feng-Hsun Chang and Meha Jain and modified by Arthur Endsley"
date: "March 30 2018"
output: 
  html_document:
    code_folding: show
    highlight: textmate
    keep_md: no
    number_sections: true
    theme: flatly
    toc: yes
    toc_float:
      collapsed: false
      smooth_scroll: true
      toc_depth: 4
---

\newcommand\expect[1]{\mathbb{E}\left[{#1}\right]}
\newcommand\var[1]{\mathrm{Var}\left[{#1}\right]}

```{r, set global theme, include=FALSE}
knitr::opts_chunk$set(fig.width=10, fig.height=8)
```

```{r, library loading, echo=FALSE, message=FALSE, warning=FALSE}
library(rmarkdown)
library(knitr) # Run R Code Chunks
library(ggplot2) # For plotting
library(dplyr)
# this is useful to convert file format, e.g. dta to csv
```


```{r}
well <- read.csv("https://raw.githubusercontent.com/OscarFHC/NRE538_GSRA/master/Labs/NRE538_GLM/wells.csv", sep = ",", header = T, comment.char = "#")
```

# Exercise 1 (R output required)     

## Question 1

---------------

Please run a logistic regression where you estimate whether a household moves based on `arsenic`, `dist`, `assoc`, and `educ`. Which variables are significantly associated with whether a household moves or not (don't interpret the variables at this stage)? 

---------------

```{r}
mod <- glm(switch ~ arsenic + dist + assoc + educ, data = well, family = binomial(link = "logit"))
summary(mod)
```

From the summary output, we see that all variables except `Intercept` and `assoc` are significantly different from zero.

## Question 2 & 3

---------------

2. Now let's do the hard part! Backtransform all of your coefficients (on `arsenic`,`dist`,`assoc`,and `educ`) and write one sentence for each interpreting your coefficient in terms of odds. Please note the unit of x when you are writing out this sentence. As a reminder, `arsenic`, `dist`, and `educ` are continuous variables and `assoc` is a categorical variable (whether a household belonged to a community association, coded as 1, or not, coded as 0).

3. Now modify the sentences above to state the % change in whether a household moves or not per unit change in each of the x variables. I know questions 2 + 3 may be challenging so please ask the GSIs for help!

---------------

Remember we are using `logit` function as the link function in this logistic regression.

$$ \text{Prob}(Y_i=1) = p_i$$  

$$ \text{logit}(p_i) = log\ (\frac{p_i}{1-p_i}) = X_i \ \beta$$  

The interpretation of the above summary table would more clear if write down the equations to describe the association between these significant variables and the probability of moving ($p$).

$$log\ (\frac{p}{1-p}) = 0 + 0.467 \cdot arsenic - 0.008 \cdot dist + 0.042 \cdot educ$$ 

We then expenitiate both sides of the equation, to get the following equation.

\begin{aligned}

\frac{p}{1-p} &= \exp(0 + 0.467 \cdot arsenic - 0.008 \cdot dist + 0.042 \cdot educ) \\
&= \exp(0) \times \exp(0.467 \cdot arsenic) \times \exp(-0.008 \cdot dist) \times \exp(0.042 \cdot educ) \\
&= 1 \times \exp(0.467 \cdot arsenic) \times \exp(-0.008 \cdot dist) \times \exp(0.042 \cdot educ)

\end{aligned}

I hope by writing down the equations, it becomes clear that

(1) When`educ`, `arsenic` and `dist` are 0, the odds ($\frac{p_i}{1-p_i}$) is 1. This means the probability of moving ($p$) is 50%. 

(2) With one unit increase of `arsenic`, the odds ($\frac{p_i}{1-p_i}$) would increase `r round(exp(summary(mod)$coefficients[2,1]), 3)` *times*. Note that it is not the probability of moving will increase `r round(exp(summary(mod)$coefficients[2,1]), 3)`. The actual probability of moving ($p$), when only increasing the `arsenic` by 1 unit, should be then calculated as followed. 

\begin{aligned}

\frac{p}{1-p} &= 1 \times \exp(0.467 \cdot 1) \times \exp(-0.008 \cdot 0) \times \exp(0.042 \cdot 0) \\
              &= 1 \times \exp(0.467 \cdot 1) \times 1 \times 1 \\
              &= 1.595 \\
p &= 1.595 \times (1-p) \\
p + 1.595\ p &= 1.595 \\
p &= \frac{1.595}{1 + 1.595} \\
p &= 0.615

\end{aligned}

When only increasing the `arsenic` by 1 unit, the probability of moving ($p$) is `r round(exp(summary(mod)$coefficients[2,1]) / (1+exp(summary(mod)$coefficients[2,1])),3)`. 

Following the same logic, one unit increase of `dist` should *decrease* the odds ($\frac{p_i}{1-p_i}$) by `r round(exp(summary(mod)$coefficients[3,1]), 3)` times and *decrease* the probability of moving ($p$) to `r round(exp(summary(mod)$coefficients[3,1]) / (1+exp(summary(mod)$coefficients[3,1])), 3)`. One unit increase of `assoc` should increase the odds ($\frac{p_i}{1-p_i}$) by `r round(exp(summary(mod)$coefficients[5,1]), 3)` times and increase the probability of moving ($p$) to `r round(exp(summary(mod)$coefficients[5,1]) / (1+exp(summary(mod)$coefficients[3,1])), 3)`. 

Note that the categorical variable (`assoc`) is not significant in this model, so that we can argue that whether any members of the household are active in community organizations does not significantly change the probability of moving (or the odds of moving). 

# Exercise 2

```{r, results='hide'}
library(faraway) # remember to install this package first!
data(gala)
head(gala)
```

The dataset contains the following variables about the species diversity on the Galapagos Islands:   

- `Species` the number of plant species found on each island    
- `Endemics` the number of endemic species found on each island    
- `Area` the area of each island ($km^2$)    
- `Elevation` the highest elevation on the island (m)   
- `Nearest` the distance to the nearest island (km)   
- `Scruz` the distance from Santa Cruz island (km)   
- `Adjacent` the area of the adjacent island (km$^2$)   

```{r,lm,results='hide'}
mod3 <- lm(Species ~ Area + Nearest + Adjacent, data = gala)
summary(mod3)
```

```{r}
mod4 <- glm(Species ~ Area + Nearest + Adjacent, data = gala, family = poisson(link = "log"))
summary(mod4)
```

## Question 1

---------------

1. Compare the results from `mod3` and `mod4` using `summary`. Which variables are significant using the linear model framework? Which variables are significant using the `glm` with Poission distribution? This suggests that making sure you get appropriate model fit is important because it can change the results of your model!    

---------------

`mod3` and `mod4` are reporting both qualitatively and quantitatively different results. In `mod3` the only significant variable is the `area` of the island, while in `mod4`, all variables are significant in determining the species number on a island. 

In addition, the from the diagnostic plots of `mod3`, we know that this model does not properly fit the data. However, you might notice that `mod4` is not perfect neither, although it slightly makes the residuals more homoskadestic and more close to normal distribution. With poisson distribution, we improved the model fit but not yet found a best model to fit our data. 


## Question 2 (EXTRA CREDIT)

---------------

For 0.5 points extra credit, interpret each of the coefficients in your Poisson model. The interpretation is NOT the same as the logistic regression because we used a different link transform. You can search online for how to interpret a possion coefficient and read this link (http://environmentalcomputing.net/interpreting-coefficients-in-glms/) to get an idea of how to back-transform the coefficient from your Poisson GLM to get a meaningful interpretation of your coefficient.   

---------------

To correctly interpret each coefficient, we again write down the equations. The link function of this model is `log` function, so that the equations describing the association between of each variable and species number (`Species`) is as follow. 

\begin{aligned}

\log(Species) &= 4.18 + 0.000432 \cdot Area + 0.00589 \cdot Nearest - 0.0000806 \cdot Adjacent\\
              
Species &= \exp(4.18 + 0.000432 \cdot Area + 0.00589 \cdot Nearest - 0.0000806 \cdot Adjacent) \\
        &= \exp(4.18) \times \exp(0.000432 \cdot Area) \times \exp(0.00589 \cdot Nearest) \times \exp(-0.0000806 \cdot Adjacent)
        
\end{aligned}

The interpretations are that

(1) When `Area`, `Nearest`, `Adjecent` are theoretically zero, the species number (`Species`) should be `r round(exp(summary(mod4)$coefficients[1,1]), 3)`.  
(2) When solely increasing island area (`Area`) by 1 km$^2$, species numner would increase `r round(exp(summary(mod4)$coefficients[2,1]), 3)` times, so that species number would become $\exp(4.18) \times \exp(0.000432 \cdot 1) = 65.398 \times 1.00043 = 65.426$. Similarly, solely increase `Nearest` by 1 unit would cause the species number to be `r round(exp(summary(mod4)$coefficients[3,1]), 3)` times higher, so that increase species number to $\exp(4.18) \times \exp(0.00589 \cdot 1) = 65.398 \times 1.005906 = 65.784$. Solely increase `Adjacent` by 1 unit would decrease species number by `r round(exp(summary(mod4)$coefficients[4,1]), 3)` times, so that species number should decrease to $\exp(4.18) \times \exp(-0.0000806 \cdot 1) = 65.398 \times 0.9999194 = 65.393$. 