---
title: "Problem Set 2"
author: "Written by Meha Jain"
date: "Week 13"
output:
  html_document:
    code_folding: show
    highlight: textmate
    keep_md: no
    number_sections: yes
    theme: flatly
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 4
  pdf_document:
    toc: yes
---

```{r, library loading, echo=FALSE, message=FALSE, warning=FALSE}
library(car)
```

Below are a series of questions related to datasets that you will pull in from our Github account. A few things:   

1. Please work on this problem set **individually** and do not ask any other students for help.    
2. Please **do not** post on Piazza for help with this problem set.  
3. This exam is **open book** so you can use any lecture notes, labs, and/or anything you find online for help!  
4. You can also ask me (Meha), Arthur, or Oscar for help.  
5. Please remember if you run a test that requires a post hoc test to answer the question, please use one. You will not get full credit for your answer if you do not use a post hoc test when needed. In the case of chi-square, we DO NOT expect you to run any post hoc tests and instead you can qualitatively explain the differences across groups by examining summary statistics.  
6. Please run all ANOVAs and ANCOVAs in the linear model framework (do not use `aov`).
7. If the assumption of homoscedasticity is ever violated, feel free to proceed with the statistical test as if the assumption was met - just make note in your answer that the assumption was violated.
8. If the assumption of normality is ever violated, we expect you to use a non-parametric test or transform your variable to meet the normality assumption. We do not expect you to use any GLMs except for a logistic regression.
9. We want you to run multi-variate regressions and logistic regressions with as many x covariates as possible.
10. When interpreting the results of a logistic regression, please backtransform your beta coefficients to make interpretations non-technical.

#  Question 1                

Let's use a dataset that has information about restaurant ratings through Yelp for restaurants from India and the United States. **Please test whether `Ratings` given to restaurants are significantly different between `Country`**.

```{r, pull in data,results='hide'}
rating = read.csv('/Users/mehajain/Desktop/yelp.csv')
head(rating)
```

The dataset contains the following variabes:      
`Restaurant`: Restaurant name      
`Country`: Country that restaurant is found in      
`City`: City the restaurant is found in        
`Price`: Price ranking of the restaurant (`1`, `2`, or `3`)     
`Rating`: Rating of the restaurant (`Average`,`Good`,`Very Good`, and `Excellent`)      

-------------------------------------------------------------------------------------
__Specific Questions__ (Show relevant R code as needed)           
-------------------------------------------------------------------------------------

a. Please write the null and alternate hypothesis (1 point).  

b. Please create a visual plot or table to answer this question (1 point).  

c. Please decide what statistical test to use and check whether your data meet the assumptions to run this test (2 points).

d. Please run the statistical test and interpret the result (1 point). Which group(s) are significantly different from one another (if any)? How are they different? 

#  Question 2                

Let's now use a dataset on a measure of poverty that is constructed by considering multiple dimensions. Specifically, this poverty index is comprised of both the number of people who are in poverty within a given country and also the intensity of poverty. If you are interested, you can read more about how the poverty index was constructed here (https://www.kaggle.com/ophi/mpi). **Please test whether `Regions` have different levels of `Poverty`.**

``` {r,load poverty data,results='hide'} 
poverty <- read.csv('/Users/mehajain/Desktop/poverty.csv')
head(poverty)
```

The dataset contains the following variables:       
`Country`: Country       
`Region`: Region country is found in        
`Poverty`: Poverty Index

-------------------------------------------------------------------------------------
__Specific Questions__ (Show relevant R code as needed)           
-------------------------------------------------------------------------------------         

a. Please write the null and alternate hypothesis (1 point).  

b. Please create a visual plot to answer this question (1 point).  

c. Please decide what statistical test to use and check whether your data meet the assumptions to run this test (2 points).

d. Please run the statistical test and interpret the result (1 point). Which group(s) are significantly different from one another (if any)? How are they different?

#  Question 3                

Let's use a real dataset from Facebook about 440 posts from 440 different people. Using this dataset, **identify which factors are significantly associated with more `shares` of a post.** 

``` {r,load rate data,results='hide'} 
fbook <- read.csv('/Users/mehajain/Desktop/facebook.csv')
head(fbook)
```

The dataset contains the following variables:   
`type`: the type of post    
`paid`: whether it was a paid advertisement or not    
`users`: the number of friends that the individual had     
`comment`: the number of people who commented on the post   
`like`: the number of people who liked the post     
`share`: the number of times that post was shared   

-------------------------------------------------------------------------------------
__Specific Questions__ (Show relevant R code as needed)           
-------------------------------------------------------------------------------------   

a. Please decide which statistical test to use and please identify which model you are going to run (2 points).

b. Please check whether your data meet the assumptions to run this test (2 points).  

c. Please run the statistical test and interpret the result. Which variables are significantly associated with the number of times a post is shared? Please write out one sentence for each significant beta coefficient explaining what it means in non-technical terms (2 points).

d. Discuss the fit of your model and whether you think it is a good or bad fit. Why? (1 point).

e. Please run a second model that tests whether the number of `likes` has a different effect on the number of `shares` depending on the `type` of post. Please interpret whether the effect of `likes` varies based on the type of post in non-technical terms (1 point).

#  Question 4                

Let's use another dataset on ratings of wine (either good or bad) and different characteristics of the wine. **Please use this dataset to identify which factors are significantly associated with ranking a wine as good (`1`) or bad (`0`).**

``` {r,load house data,results='hide'} 
wine <- read.csv('/Users/mehajain/Desktop/wine.csv')
head(wine)
```

The dataset contains the following variables:   
`citricacid`: the amount of citric acid per glass   
`sugar`: the amount of sugar per glass   
`chlorides`: the amount of chlorides per glass   
`density`: the density of the wine   
`sulphates`: the amount of sulphates per glass    
`alcohol`: the alcohol content per glass     
`quality`: whether the wine is good (`1`) or bad (`0`)

-----------------------------------------------------------------------------------------------
__Specific Questions__ (Show relevant R code as needed)           
-----------------------------------------------------------------------------------------------

a. Please decide which statistical test to use and please identify which model you are going to run (2 points).

b. Please run the statistical test and interpret the result. Which variables are significantly associated with whether a wine is ranked as good or bad (1 point).

c. For all significant variables, please interpret each beta coefficient in non-technical terms. Since the x variables are unitless, you can simply say 'for every 1 unit change in x...' (2 points).

#  Question 5
For 3 points extra credit, please find your own dataset and run a multiple linear regression. For full credit we will expect that you will check for multicollinearity, check all assumptions, and interpret results in non-technical terms. 